Parsing Postfix log files

### 2008-10-28 doc/logparser-implementation.tex
First half of the corrections to section 5.

\newpage
\section{Postfix Parser Implementation}

@@

The algorithm described and the program implementing it make a small number
of (hopefully safe and reasonable) assumptions:

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time: there are 375,750
lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
code.} there are advantages to treating Postfix as a black box during
parser development:

@@

@@

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  A separate project
        could compare the empirical knowledge inherent in the parsing
        algorithm with the documentation and source code of Postfix to see
        how closely the two agree.

@@

@@

The database schema can be conceptually divided in two: the rules, which
are used to parse log files, and the data saved from the parsing of log
files.  Rules have the fields required to parse the log lines, extract data
to be saved, and the action to be executed; they also have several fields
that aid the user in understanding the meaning of the log lines parsed by
each rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

@@

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rule that produced the result is
available for inspection --- each result references the rule that created
it.  There is no ambiguity about which rule resulted in a particular
result, eliminating one potential source of confusion.

@@

@@

    \item [id] A unique identifier for each rule that other tables can use
        to refer to a specific rule.

@@

@@

            \item [BOUNCED] The mail has bounced, because of a mail loop,
                delivery failure, or five day timeout.

@@

            \item [DISCARDED] Postfix discarded the mail it was in the
                process of accepting, because it was either larger than the
                limit set by the administrator, or the client timed out or
                disconnected.

            \item [EXPIRED] The mail has been in the queue for five
                days\footnote{As with much of Postfix's behaviour, this is
                the default value but can be changed by the administrator
                if they choose.} without successful delivery.  A bounce
                mail will be generated and sent to the sender address.

@@

    \item [regex] The \regex{} to match the log line against.  The \regex{}
        will first have several keywords expanded: this simplifies reading
        and writing rules; avoids needless repetition of complex \regex{}
        components; allows the components to be corrected and/or improved
        in one location; and makes each \regex{} largely self-documenting.

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Efficiency concerns are
        discussed in \sectionref{parser efficiency}, with caching compiled
        \regexes{} covered in \sectionref{Caching compiled regexes}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline{} \tab{}
        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline{} i.e.\
        semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching capture from the
        \regex{} on the right hand side.  Data specified by
        \texttt{result\_cols} will be entered in the \texttt{results}
        table, whereas \texttt{connection\_cols} populates the
        \texttt{connection} table.  The list of acceptable variable names
        is:

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline{} \tab{} server\_hostname} and \texttt{helo}.\newline{}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline{} \tab{} size,} and \texttt{data}

        Additionally \texttt{child} and \texttt{pid} are accepted and used
        respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
        in the database.

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data that is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This feature is intended for use with small
        amounts of data rather than large amounts in any one rule, so
        dealing with escape sequences was deemed unnecessary.} is permitted
        on the right hand side of the assignment.

@@

    \item [queueid] Specifies the capturing field in the \regex{}
        containing the queueid, or zero if the log line does not contain a
        queueid.  Many log lines will not contain a queueid, e.g.\
        rejections logged before a mail has been accepted (a queueid will
        not have been allocated), or log lines that are not tied to one
        particular mail.

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully matches.  At the start of each run
        the program sorts the rules in descending order of hits, and at the
        end of the run updates every rule's hits field in the database.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, reducing the program's execution
        time.  Rule ordering for efficiency is discussed in
        \sectionref{rule ordering for efficiency}.

@@

    \item [cluster\_group] A reference to the \texttt{cluster\_group}
        table.  That table is used by the Decision Tree algorithm described
        in a separate document.

@@

@@

XXX REWRITE THE SENTENCE AND FOOT NOTE\@.

Every interesting\footnote{All log lines are interesting unless the
postfix\_action of the rule that parses them is \texttt{INFO} or
\texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
will have an entry in the results table, e.g.\ rejecting an \SMTP{}
command, delivering a mail, or bouncing a mail (see
\sectionref{postfix_action} for typical values of
\texttt{postfix\_action}).  Each row is associated with a single
connection, though there may be many results per connection.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection.  This field should be
        ignored if the result is not a rejection, i.e.\ the action field of
        the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present for a rejection or final
        delivery; results initially missing an \SMTP{} code will duplicate
        the \SMTP{} code of other results in the connection.  Some final
        delivery log lines do not contain an \SMTP{} code: in those cases
        the code is specified by the rule, based on the success or failure
        represented by the log line.

@@

    \item [recipient] The recipient address; there may be multiple
        recipient addresses per mail, but only one per result.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an \DNSBL{}\@.

    \item [timestamp] The time the log line was logged, in seconds since
        the epoch.

XXX DO I NEED A SUMMARY HERE\@?

The table containing the rules used by the parser and both tables
containing the data extracted from the Postfix log files were described,
with the purpose of each field discussed in detail.  A clear,
comprehensible schema is essential when using the extracted data; it is
more important when using the data than when storing it, because storing
the data is a write-once operation, whereas utilising the data requires
frequent searching, sorting and manipulation of the data to produce
customised reports or statistics.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the appropriate \DNS{} entries are
missing, i.e.\ mail could not be delivered to the sender's address (for
full details see~\cite{reject-unknown-sender-domain}).

The example rule below matches the following log line:

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
  <foo@example.com>: Sender address rejected: Domain not found;
  from=<foo@example.com> to=<info@example.net>
  proto=SMTP helo=<smtp.example.com>
\end{verbatim}

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & \daemon{smtpd}                                    \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\
cluster\_group      & 400                                               \\

@@

Additional data will be captured automatically when the \regex{} contains 
\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
connection\_cols start at 5.  The various fields are used as follows;

@@

    \item [name, description, restriction\_name and postfix\_action:] are
        not \newline{} used by the algorithm, they serve to document the rule
        for the user's benefit.

    \item [program and regex:] If the program in the rule equals the
        program that logged the log line, a match using the rule's
        \regex{} will be attempted against the log line; if the match is
        successful the action will be executed, if not the next rule will
        be tried.  If the program-specific rules do not match the log line,
        the parser will fall back to generic rules; if those rules are
        unsuccessful a warning will be issued.

    \item [action:] The action to be executed if the \regex{} matches
        successfully (see \sectionref{actions-in-detail} for full details).

    \item [result\_cols, connection\_cols, result\_data and
        connection\_data:] are \newline{} used by the action to extract and
        save data matched by the \regex{}.

    \item [queueid:] The index of the capture in the \regex{} that
        supplies the queueid, or zero if the log line does not contain a
        queueid.  This allows the correct mail to be found by queueid and
        actions performed on it.

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run; hits\_total is the sum of hits over every parsing run,
        but is otherwise unused by the algorithm.

    \item [cluster\_group] The cluster\_group attribute is used by the
        Decision Tree algorithm described in a separate document; the
        parser does not use it in any way.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \DNSBL{}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial, and also occurs much more
frequently.  The implementation provides a program to ease the process of
creating new rules from unparsed log lines, based on the algorithm
developed by Risto Vaarandi~\cite{risto-vaarandi} for his
\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
be outlined as part of the general explanation below.

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is variable or part of the fixed pattern.  This classification can be
used to group similar log lines and generate a \regex{} to match each group
of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the log input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        that the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each log line, incrementing a counter for each hash.  The
        counters will be used in the next pass to filter out words: if the
        word's hash does not have a high frequency, the word itself cannot
        have a high frequency, and there is no need to maintain a counter
        for it, reducing the number of counters and thus the program's
        memory consumption.

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a common word does not indicate similarity
        between log lines unless it occupies the same position within both
        log lines.\footnote{If a variable term within a line contains
        spaces, it will appear to the algorithm as two words rather than
        one.  This will alter the position of subsequent words, so a word
        occurring in different positions in two log lines \textit{may\/}
        indicate similarity, but the algorithm does not attempt to deal
        with this possibility.}  The algorithm maintains a counter for each
        \textit{(word, word's position within the log line)\/} tuple,
        incrementing it each time that word occurs in that position.

        The original algorithm only maintains counters for words whose hash
        result from the previous step has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.

        As time goes on the amount of memory typically available increases
        and the requirement to reduce memory requirements decreases, hence
        the modified algorithm omits the hashing step and maintains
        counters for all tokens. In addition most uncommon or infrequently
        occurring words will have been substituted with keywords during the
        first step, vastly reducing the number of tuples to maintain
        counters for.

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the log line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters added
        together.  If the \regex{}'s counter is lower than the threshold
        supplied by the user the \regex{} is discarded; this second
        threshold is independent of the threshold used to differentiate
        between fixed and variable words, but once again 1\% of input log
        lines is a good starting point; obviously the threshold depends on
        the number and content of input log lines.  The new \regexes{} are
        displayed for the user to add to the database, either as new rules
        or merged into the \regexes{} of existing rules; the counter for
        each \regex{} is also displayed, giving the user an indication of
        how many of the input log lines that \regex{} should match.
        Discarding \regexes{} will result in some of the input log lines
        not being matched; the utility should be run again once the
        unmatched log lines have been isolated by parsing the original log
        files once the parser has been augmented with the new rules.

@@

@@

\subsection{Parsing Algorithm}

\label{parsing-algorithm}

EMPHASISE THAT THE LOGS CAUSE THE INTERDEPENDENCIES, NOT THE
ARCHITECTURE\@.

\subsubsection{Introduction}

XXX MERGE WITH PREVIOUS PARAGRAPHS\@.

This section covers the following topics:

    \item A high level overview of the algorithm.

    \item The first set of complications encountered: initially obvious
        difficulties that had to be overcome.

    \item A flow chart showing common paths a mail can take through the
        algorithm, a discussion of the emergent behaviour observed in the
        parser, and an explanation of the paths shown in the flow chart.

    \item A brief description of the framework.

    \item The actions the parser makes available to rules are covered in
        detail.

    \item Additional complications that have arisen during the development
        of this parser are documented, the solutions to those
        complications, and where those solutions are implemented in the
        algorithm

\subsubsection{A high level overview}

@@

Unfortunately both high level views ignore the many complications
encountered.

\subsubsection{Flow chart}

XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DO NOT THINK THAT
RE-INJECTION IS IMPORTANT ENOUGH TO WARRANT IT\@.  THE EXPLANATION OF
TRACKING (\textsection{}5.4.3) COULD BE MOVED TO ADDITIONAL
COMPLICATIONS\@.  POSSIBLY MERGE MAIL ENTRY AND DELIVERY IN HERE\@;
TRACKING SHOULD BE IN COMPLICATIONS --- MERGE WITH RE-INJECTED MAILS\@.

\Figureref{flow chart image part 1} and \figureref{flow chart image part 2}
show the paths the data representing a mail can take through the parser
algorithm.  The flow chart covers the most common paths only; there are
additional, uncommon paths excluded for the sake of clarity; details of the
deviations can be found in \sectionref{complications}.

\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
1}{flow chart image part 1}

\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
2}{flow chart image part 2}

@@

The actions introduce some of the complexity seen in the flow chart: all
explicit branches and the transitions between some stages (19--20,
21--22--23--24) in the second half of the flow chart
(\figureref{flow chart image part 2}) are encoded within the
actions; the transitions between the other stages (16--17, 24--25) in the
second half arise from the interaction between actions

However the transitions between stages in the first half of the flow chart
(\figureref{flow chart image part 1}) are not encoded anywhere in
the parser; the transitions are determined by the ordering of log lines in
the log file.  The complexity of the first 12 stages in the flow chart
emerges from the interaction of simple rules and actions, easing the
process of adding new actions (described in \sectionref{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart.\footnote{The solution to the
complication in \sectionref{out of order log lines} (out of order log
lines) requires a list of the acceptable combinations of Postfix programs a
mail can pass through; however this does not correspond to the
\textit{parser actions\/} a mail must pass through.  Before that
complication was overcome, i.e.\ for the first half of the development of
this parser, there was nothing in the parser explicitly encoding the states
or paths a mail could take; no other component in the parser has any need
for this flow of data to be specified.}

@@

    \item The program will not detect parsing the same log file twice,
        resulting in the database containing duplicate entries.

>>>>

\section{Postfix Parser Implementation}

XXX WRITE AN INTRODUCTION\@.

XXX UPDATE REFERENCES WHEN ARCHITECTURE AND IMPLEMENTATION ARE MORE FLESHED
OUT\@.

XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, ETC\@.

XXX THIS MIGHT BE USEFUL SOMEWHERE\@:

@@

\parsername{} makes a small number of (hopefully safe and reasonable)
assumptions:

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  There are advantages to
treating Postfix as a black box during parser development:

@@

    \item Reading and understanding the source code would require a
        significant investment of time: Postfix 2.5.5 has 17MB of source
        code.

@@

    \item The parser acts as a second source of information about Postfix's
        operation, using information gathered from empirical evidence.  A
        separate project could compare the empirical knowledge inherent in
        the parsing algorithm with the documentation and source code of
        Postfix to see how closely the two agree.

@@

XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.

@@

The database schema can be conceptually divided in two: the rules used to
parse log files, and the data saved from the parsing of log files.  Rules
have the fields required to parse the log lines, extract data to be saved,
and specify the action to be executed; they also have several fields that
aid the user in understanding the meaning of the log lines parsed by each
rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

@@

An important but easily overlooked benefit of storing the rules in the
database is the link between rules and results: if more information is
required when examining a result, the rule that produced the result is
available for inspection: each result references the rule that created it.
There is no ambiguity about which rule resulted in a particular result,
eliminating one potential source of confusion.

A clear, comprehensible schema is essential when using the extracted data;
it is more important when using the data than when storing it, because
storing the data is a write-once operation, whereas utilising the data
requires frequent searching, sorting and manipulation of the data to
produce customised reports or statistics.

@@

XXX SHOULD I USE TEXTTT OR SIMILAR ON FIELD NAMES\@?

@@

\begin{eqlist}

    \item [id] A unique identifier for each rule that other tables can use
        when referring to a specific rule.

@@

        XXX SHOULD THIS LIST BE HERE\@?  MAYBE MOVE TO AFTER THE FIELD
        LIST\@?  DO I NEED THIS LIST AT ALL\@?

@@

            \item [BOUNCED] The mail has bounced, because of a mail loop,
                delivery failure, or five day timeout.\footnote{As with
                much of Postfix's behaviour, five days is the default value
                but can be changed by the administrator if they choose.}

@@

            \item [DISCARDED] Postfix discarded the mail it was in the
                process of accepting, because it was either larger than the
                size limit set by the administrator, or the client timed
                out or disconnected.

            \item [EXPIRED] The mail has been in the queue for five
                days\footnotemark[\value{footnote}]XXX CHECK IF THIS
                FOOTNOTE MAKES SENSE without successful delivery.  A bounce
                mail will be generated and sent to the sender address.

@@

    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
        EXTRACTION IS FINISHED\@.  The \regex{} to match the log line
        against.  The \regex{} will first have several keywords expanded:
        this simplifies reading and writing rules; avoids needless
        repetition of complex \regex{} components; allows the components to
        be corrected and/or improved in one location; and makes each
        \regex{} largely self-documenting.  For efficiency the keywords are
        expanded and every rule's \regex{} is compiled before attempting to
        parse the log file --- otherwise each \regex{} would be recompiled
        each time it was used, resulting in a large, data dependent
        slowdown.  Efficiency concerns are discussed in \sectionref{parser
        efficiency}, with caching compiled \regexes{} covered in
        \sectionref{Caching compiled regexes}.

    \item [result\_data] Sometimes rules need to save a result that is not
        present in the log line: e.g.\ setting \texttt{smtp\_code} when
        mail is accepted.  The format is: \newline{} \tab{} \texttt{XXX
        UPDATE EXAMPLE\@; smtp\_code = 550; recipient = 2, sender = 4;}
        \newline{} i.e.\ semi-colon or comma separated assignment
        statements, with the variable name on the left and the value on the
        right hand side.\footnote{Commas and semi-colons cannot be escaped
        and thus cannot be used.  This feature is intended for use with
        small amounts of data rather than large amounts in any one rule, so
        dealing with escape sequences was deemed unnecessary.}  Any field
        in the results table can be set in this way; additionally
        \texttt{child} and \texttt{pid} are accepted and used respectively
        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
        actions, but do not cause data to be entered in the database.

    \item [connection\_data] The connections table equivalent of
        \texttt{result\_data}.

@@

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully matches.  At the start of each run
        the parser sorts the rules in descending order of hits, and at the
        end of the run it updates every rule's hits field in the database.
        Assuming that the distribution of log lines is reasonably
        consistent across log files, rules matching more commonly occurring
        log lines will be tried before rules matching less commonly
        occurring log lines, reducing the parser's execution time.  Rule
        ordering for efficiency is discussed in \sectionref{rule ordering
        for efficiency}.

@@

%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
%        table.  That table is used by the Decision Tree algorithm described
%        in a separate document.

\end{eqlist}

@@

\begin{eqlist}

@@

Every log line a row in the results table, unless the
\texttt{postfix\_action} field of the rule which parsed it was INFO or
IGNORE\@.  Each row in the results table is associated with a single
connection, and there may be many results per connection.

\begin{eqlist}

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        mechanism that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection.  This field should be
        ignored if the result is not a rejection, i.e.\ the action field of
        the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present in a log line for a
        rejection or final delivery; results initially missing an \SMTP{}
        code will duplicate the \SMTP{} code of other results in that
        connection.  Some final delivery log lines do not contain an
        \SMTP{} code: in those cases the code is specified by the rule,
        based on the success or failure represented by the log line.

@@

    \item [recipient] The recipient address; there may be multiple
        recipient addresses per mail or connection.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \DNSBL{}\@.

    \item [timestamp] The time at which the log line was logged, in seconds
        since the epoch.

\end{eqlist}

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the domain in the sender address does
not have an A, AAAA, or MX \DNS{} entry, i.e.\ mail could not be delivered
to the sender's address (for full details
see~\cite{reject-unknown-sender-domain}).

Example log line matched by this rule:

% RFC 3330 says that 192.0.2.0/24 is reserved for example use.

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550
  <foo@example.com>: Sender address rejected: Domain not found;
  from=<foo@example.com> to=<info@example.net>
  proto=SMTP helo=<smtp.example.com>
\end{verbatim}

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & \daemon{smtpd}                                    \\
XXX                 & UPDATE REGEX WHEN AUTOMATIC SAVING IS FINISHED    \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & DELIVERY\_REJECTED                                \\
hits                & 0                                                 \\
hits\_total         & 0                                                 \\
priority            & 0                                                 \\
%cluster\_group      & 400                                               \\

@@

Fields matched by the regex will be automatically saved to the results and
connections tables.

@@

    \item [name, description, restriction\_name, and postfix\_action:] are
        not \newline{} used by the parser, they serve to document the rule
        for the user's benefit.

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's \regex{} will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

    \item [action:] The action to be executed if the \regex{} matches
        successfully.  See \sectionref{actions-in-detail} for details of
        the actions implemented by the parser, \sectionref{actions in
        architecture} for the role of actions in the parser architecture.

    \item [result\_data and connection\_data:] are used to provide XXX

    \item [hits, hits\_total, and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run; hits\_total is the sum of hits over every parsing run,
        but is otherwise unused by the parser.

%    \item [cluster\_group] The cluster\_group attribute is used by the
%        Decision Tree algorithm described in a separate document; the
%        parser does not use it in any way.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \DNSBL{}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions can be difficult, but adding new rules
to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \SLCT{}~\cite{slct-paper}.  The differences between the
two algorithms will be outlined as part of the general explanation below.

The core of the \SLCT{} algorithm is quite simple: log lines are generally
created by substituting variable words into a fixed pattern, and analysis
of the frequency with which each word occurs can be used to determine
whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a
\regex{} to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        that the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        \regexes{}; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each log line, incrementing a counter for each hash.  The
        counters will be used in the next pass to filter out words: if the
        word's hash does not have a high frequency, the word itself cannot
        have a high frequency, so there is no need to maintain a counter
        for it, reducing the number of counters required and thus the
        program's memory consumption.

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a word occurring in two lines does not indicate
        similarity unless it occupies the same position within both log
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two or more words rather than
        one.  This will alter the position of subsequent words, so a word
        occurring in different positions in two log lines \textit{may\/}
        indicate similarity, but the algorithm does not attempt to deal
        with this possibility.}  The algorithm maintains a counter for each
        \texttt{(word, word's position within the log line)} tuple,
        incrementing it each time that word occurs in that position.

        The original algorithm only maintains counters for words whose hash
        result from the previous step has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm at a cost of increased CPU
        usage.

        As time goes on the amount of memory typically available increases
        and the requirement to reduce memory requirements decreases, hence
        the modified algorithm omits the hashing step and maintains
        counters for all tokens. In addition most of the infrequently
        occurring words will have been substituted with keywords during the
        first step, vastly reducing the number of tuples to maintain
        counters for.

    \item [Classify words based on their frequency]  The frequency of each
        \texttt{(word, word's position within the log line)} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.  XXX SHOULD I USE \verb![\S]+! INSTEAD\@?

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter is maintained for each
        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
        resulting duplicate \regexes{} are combined, and their counters
        added together.  If a \regex{}'s counter is lower than the
        threshold supplied by the user the \regex{} is discarded; this
        second threshold is independent of the threshold used to
        differentiate between fixed and variable words, but once again 1\%
        of log lines is a good starting point.  The new \regexes{} are
        displayed for the user to add to the database, either as new rules
        or merged into the \regexes{} of existing rules; the counter for
        each \regex{} is also displayed, giving the user an indication of
        how many of the log lines that \regex{} should match.  Discarding
        \regexes{} will result in some of the log lines not being matched;
        when the rules have been augmented with the new \regexes{}, the
        original log files should be parsed again, and any remaining
        unparsed log lines used as input to this utility.

@@

XXX MERGE logs2regexes WITH check-regexes.

@@

\subsection{Parser Flow Chart}

XXX MIGHT NEED A SMALL INTRO\@.

XXX MOVE TO AFTER ACTIONS\@.

@@

\Figureref{flow chart image} shows the paths the data representing a mail
can take through the parser algorithm.  The flow chart covers the most
common paths only, the complications described in
\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
ARE THE PATHS DESCRIBED\@?

\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
image}

@@

The actions introduce some of the complexity seen in the flow chart: all
explicit branches and the transitions between some stages (19--20,
21--22--23--24) in the second half of the flow chart are encoded within the
actions; the transitions between the other stages (16--17, 24--25) in the
second half arise from the interaction between actions

However the transitions between stages in the first half of the flow chart
(\figureref{flow chart image}) are not encoded anywhere in
the parser; the transitions are determined by the ordering of log lines in
the log file.  The complexity of the first 12 stages in the flow chart
emerges from the interaction of simple rules and actions, easing the
process of adding new actions (described in \sectionref{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart.\footnote{The solution to the
complication in \sectionref{out of order log lines} (out of order log
lines) requires a list of the acceptable combinations of Postfix programs a
mail can pass through; however this does not correspond to the
\textit{parser actions\/} a mail must pass through.  Before that
complication was overcome, i.e.\ for the first half of the development of
this parser, there was nothing in the parser explicitly encoding the states
or paths a mail could take; no other component in the parser has any need
for this flow of data to be specified.}

@@

    \item The parser will not detect parsing the same log file twice,
        resulting in the database containing duplicate entries.

### 2008-10-28 doc/logparser-introduction.tex
First half of the corrections to section 5.

Section~\ref{parsing-algorithm} contains the core of the paper, describing
a naive parsing algorithm and the complications initially encountered that
shaped the full algorithm.  A flow chart and a discussion of the emergent
behaviour exhibited by the algorithm accompanies a comprehensive
explanation of the different stages of the initial algorithm.  The
framework that actions and rules fit into is documented, then the actions
taken during execution of the algorithm are described, followed by the
process of adding a new action.  The section concludes with an in-depth
description of the further complications discovered, and their solutions
that complete the parser.

>>>>

Section~XXX contains the core of the paper, describing
a naive parsing algorithm and the complications initially encountered that
shaped the full algorithm.  A flow chart and a discussion of the emergent
behaviour exhibited by the algorithm accompanies a comprehensive
explanation of the different stages of the initial algorithm.  The
framework that actions and rules fit into is documented, then the actions
taken during execution of the algorithm are described, followed by the
process of adding a new action.  The section concludes with an in-depth
description of the further complications discovered, and their solutions
that complete the parser.
